Embracing the Laws of Physics: \\ Three Reversible Models of Computation

### 2018-06-15 survey/survey.tex
fix intro of section 4

\noindent Why \emph{quasi}? The reasons are beyond our scope, but the
interested reader can read Sec.~$2.4$ and Ch.~$4$ in the HoTT
book~\cite{hottbook}.  There are several conceptually different, but
equivalent, ``better'' definitions.  We record just one here:

@@

In the previous sections, we examined equivalences between
conventional data structures, i.e, sets of values and structured trees
of values. We now consider a richer, less conventional, but
foundational notion of data: programs themselves. Indeed, universal
computation models fundamentally rely on the fact that \emph{programs
are (or can be encoded as) data}, e.g., a Turing machine can be
encoded as a string that another Turing machine (or even the same
machine) can manipulate. In our setting, the programs developed in the
previous section are reversible deformations between structured finite
types. We now ask whether these programs can themselves
be subject to (higher-level) reversible deformations?

@@

\begin{center}
\begin{tikzpicture}[scale=0.7,every node/.style={scale=0.8}]
  \draw[>=latex,<->,double,red,thick] (2.25,-1.2) -- (2.25,-2.9) ;
%%  \node at (3.3,-2) {$\mathit{swapl}_{+\Leftrightarrow}$} ;
%%  \node at (2.5,-1.3) {$((c_2~\oplus~c_1)~\odot~\mathit{swap}_{+})$};
%%  \node at (2.5,-2.7) {$(\mathit{swap}_{+}~\odot~(c_1~\oplus~c_2))$};
  \draw (-2,-2) ellipse (0.5cm and 1cm);
  \draw[fill] (-2,-1.5) circle [radius=0.025];
  \node[below] at (-2.1,-1.5) {$A$};
  \draw[fill] (-2,-2.5) circle [radius=0.025];
  \node[below] at (-2.1,-2.5) {$B$};

@@

  \draw     (2.5,-3)  -- (3.5,-3) -- (3.5,-4) -- (2.5,-4) -- cycle ;
  \draw     (2.5,-4)  -- (3.5,-4) -- (3.5,-5) -- (2.5,-5) -- cycle ;

  \draw     (1,1)  -- (2,1) -- (2,0) -- (1,0) -- cycle ;
  \draw     (1,0)  -- (2,0) -- (2,-1) -- (1,-1) -- cycle ;

@@

\end{tikzpicture}
\end{center}
The top path is the $\Pi$ program
$(c_1~\oplus~c_2)~\odot~\swapp$ which deforms the
type $A$ by $c_1$, deforms the type $B$ by $c_2$, and deforms the
resulting space by a twist that exchanges the two injections into the
sum type. The bottom path performs the twist first and then deforms
the type $A$ by $c_1$ and the type $B$ by $c_2$ as before. If one
could imagine the paths are physical wires and the deformations $c_1$
and $c_2$ as arbitrary deformations on these wires then, holding the
points $A$, $B$, $C$, and $D$ fixed, it is possible to rotate the top
part of the diagram to become identical to the bottom one. That
rotation can be undone (reversed), which takes the bottom
part of the diagram into the top part.  In other
words, there exists a deformation of the program
$(c_1~\oplus~c_2)~\odot~\swapp$ to the program
$\swapp \odot (c_2~\oplus~c_1)$. We can also show that this
means that, as permutations, $(c_1~\oplus~c_2)~\odot~\swapp$ and
$\swapp \odot (c_2~\oplus~c_1)$ are equal. And, of course, not
all programs between the same types can be deformed into one
another. The simplest example of inequivalent deformations
are the two automorphisms of $1+1$, namely $\idc$ and $\swapp$.

>>>>

\noindent Why \emph{quasi}? The reasons are beyond our scope, but the
interested reader can read Sec.~$2.4$ and Ch.~$4$ in the
Homotopy Type Theory (HoTT) book~\cite{hottbook}.
There are several conceptually different, but
equivalent, ``better'' definitions.  We record just one here:

@@

In the previous sections, we examined equivalences between
conventional data structures, i.e., sets of values and structured trees
of values. We now consider a richer but
foundational notion of data: programs themselves. Indeed, universal
computation models crucially rely on the fact that \emph{programs
are (or can be encoded as) data}, e.g., a Turing machine can be
encoded as a string that another Turing machine (or even the same
machine) can manipulate. Similarly, first-class functions are
the \emph{only} values in the $\lambda$-calculus.
In our setting, the programs developed in the
previous section are reversible deformations between structured finite
types. We now ask whether these programs can themselves
be subject to (higher-level) reversible deformations?

@@

\jc{we need to make sure that the pictures we draw in the
previous section are consistent with this one, otherwise we'll
completely confuse our readers. Or we redraw this one to be
consistent.}
\begin{center}
\begin{tikzpicture}[scale=0.7,every node/.style={scale=0.8}]
  \draw[>=latex,<->,double,red,thick] (2.25,-1.2) -- (2.25,-2.9) ;
%%  \node at (3.3,-2) {$\mathit{swapl}_{+\Leftrightarrow}$} ;
%%  \node at (2.5,-1.3) {$((c_2~\oplus~c_1)~\odot~\mathit{swap}_{+})$};
%%  \node at (2.5,-2.7) {$(\mathit{swap}_{+}~\odot~(c_1~\oplus~c_2))$};
  \draw (-2,-2) ellipse (0.5cm and 1cm);
  \draw[fill] (-2,-1.5) circle [radius=0.025];
  \node[below] at (-2.1,-1.5) {$A$};
  \draw[fill] (-2,-2.5) circle [radius=0.025];
  \node[below] at (-2.1,-2.5) {$B$};

@@

  \draw     (2.5,-3.1)  -- (3.5,-3.1) -- (3.5,-3.9) -- (2.5,-3.9) -- cycle ;
  \draw     (2.5,-4.1)  -- (3.5,-4.1) -- (3.5,-4.9) -- (2.5,-4.9) -- cycle ;

  \draw     (1,0.9)  -- (2,0.9) -- (2,0.1) -- (1,0.1) -- cycle ;
  \draw     (1,-0.1)  -- (2,-0.1) -- (2,-0.9) -- (1,-0.9) -- cycle ;

@@

\end{tikzpicture}
\end{center}
The top path is the $\Pi$ program
$(c_1~\oplus~c_2)~\odot~\swapp$ which deforms the
type $A$ by $c_1$, deforms the type $B$ by $c_2$, and deforms the
resulting space by a twist that exchanges the two injections into the
sum type. The bottom path performs the twist first and then deforms
the type $A$ by $c_1$ and the type $B$ by $c_2$ as before. One
could imagine the paths are physical \emph{elastic} wires in $3$ space, where
the deformations $c_1$ and $c_2$ as arbitrary deformations on these wires, and
the twists do not touch but are in fact well-separated. Then, holding the
points $A$, $B$, $C$, and $D$ fixed, it is possible to imagine
sliding $c_1$ and $c_2$ from the top wire rightward past the
twist, and then using the elasticity of the wires, pull the
twist back to line up with that of the bottom --- thus making
both parts of the diagram identical.  Each of these moves
can be undone (reversed), and doing so would take the bottom
part of the diagram into the top part.  In other
words, there exists a deformation of the program
$(c_1~\oplus~c_2)~\odot~\swapp$ to the program
$\swapp \odot (c_2~\oplus~c_1)$. We can also show that this
means that, as permutations, $(c_1~\oplus~c_2)~\odot~\swapp$ and
$\swapp \odot (c_2~\oplus~c_1)$ are equal. And, of course, not
all programs between the same types can be deformed into one
another. The simplest example of inequivalent deformations
are the two automorphisms of $1+1$, namely $\idc$ and $\swapp$.
